---
title: "Assignment 1 Task 1"
author: "Michelle D. Mohr"
format: 
  html: 
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

# Task 1

Dataset Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3

## Data wrangling and visualization (Sierra amphibians)

1)  Read in .xlsx file and do some data wrangling
2)  Create two data visualizations and put them together in a finalized compound figure

### Task 1 Part 1

A graph of total mountain yellow-legged frog (Rana muscosa) count each year across all water bodies, by life stage excluding the ‘EggMass’ level.

Read in and wrangle the amphibian data
```{r read in and wrangle the data}
library(tidyverse)
library(here)
#install.packages("readxl")
library(readxl)
## read in the excel file ##
sierra_amphibians <- read_excel("sierra_amphibians.xlsx")

## data wrangling based on deliverables 
### deliverable 1) graph of total mountain yellow- legged frog count each year across all water bodies, by life stage excluding the "EggMass' level 
### deliverable 2) a column graph containing total counts (over all years of study) of combined adult and subadult endangered mountain yellow-legged frogs observed in the five lakes with the greatest total observed counts

# first convert the date column to standard date 
#install.packages("lubridate")
library(lubridate)

#change the survey_date column to a new column called date that is recognized as a date
#extract the year
#filter for RAMU
sierra_amphibians_mylf <- sierra_amphibians %>%
  mutate(date = lubridate::ymd(survey_date)) %>% 
  mutate(year = lubridate::year(date)) %>%
  filter(amphibian_species == 'RAMU')
  
#select specific columns to clean dataframe
#group by life stage to be able to plot
sierra_amphibians_mylf_count <- sierra_amphibians_mylf %>%
  select(-survey_date) %>%
    group_by(amphibian_life_stage)

#take out EggMass life stage 
mylf_count <- sierra_amphibians_mylf_count %>%
  filter(amphibian_life_stage != "EggMass") %>%
    group_by(amphibian_life_stage)

#summarize counts per year by life stage
summary_per_stage <- mylf_count %>%
  group_by(year, amphibian_life_stage) %>%
  summarize(total_amphibian_number = sum(amphibian_number))
```

Create a lifestage plot (excluding EggMass)
```{r lifestage plot}
lifestage_plot <- ggplot(data = summary_per_stage, aes(x = year, y = total_amphibian_number, fill = amphibian_life_stage)) + 
  geom_bar(stat = "identity", position = "dodge") +  # Use geom_bar() with stat = "identity" to plot y values directly
  theme_light() +
  labs(
    x = "Year",
    y = "Count",
    title = "Mountain Yellow Legged Frog Counts per Year",
    caption = "Mountain Yellow Legged Frog (RAMU) counts per year by life stage"
  ) +
  scale_fill_discrete(name = "Amphibian Life Stage",
                      labels = c("Adult", "Sub Adult", "Tadpole"))
lifestage_plot
```

### Task 1 Part 2

A column graph containing total counts (over all years of the study) of combined adult and subadult endangered mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts.

```{r lake plot}
mylf_lakes <- sierra_amphibians_mylf_count %>%
  filter(amphibian_life_stage != "Tadpole") %>%
  group_by(lake_id)%>%
  summarize(total_amphibian_number = sum(amphibian_number)) %>% 
  slice_max(total_amphibian_number, n=5)

mylf_lakes <- mylf_lakes %>%
  mutate(lake_id = recode(lake_id,
                          "50183" = "Lake 50183",
                          "10223" = "Lake 10223",
                          "10329" = "Lake 10329",
                          "20196" = "Lake 20196",
                          "10225" = "Lake 10225"))

mylf_lakes <- mylf_lakes %>%
  mutate(lake_id = as.character(lake_id))

lakeplot <- ggplot(data = mylf_lakes, aes(x = fct_reorder(lake_id, -total_amphibian_number),
                              y = total_amphibian_number)) +
  geom_col() + 
  labs(x = "Lake", y = "Count",
       caption = "Top five populations of combined adult and sub adult \nMountain Yellow Legged Frogs (RAMU) over all years of the study") +
  ggtitle("Top 5 Lakes by Mountain Yellow Legged Frog Population") +
  theme_light() 

lakeplot
```
Combine the lifestage plot and the lake plot
```{r patch the two plots together}
library(patchwork)
lifestage_lake <- lifestage_plot + lakeplot 
lifestage_lake + 
  plot_annotation(
    caption = "Two plots showing population of Mountain Yellow Legged Frog by life stage over all years of the study and the lakes with the five highest populations.")
```
