---
title: "Assignment 1 Task 2"
author: "Michelle D. Mohr"
format: 
  html: 
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

# Task 2

Dataset Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3

## Willamette Falls fish passage time series summary 

1) Original time series
2) Seasonplots
3) Annual counts by species

### Task 2 Part 1: Original time series

A static graph of adult passage for coho, jack coho, and steelhead salmon.

```{r read in data and wrangle}
library(tidyverse)
library(here)
library(tsibble)
library(feasts)
library(fable)
library(lubridate)
willfalls <- read.csv('willamette_fish_passage.csv') %>%
  janitor::clean_names() %>%
  mutate(date = lubridate::ydm(date))
  

willfalls_ts <- willfalls %>%
  select(date, steelhead, coho, jack_coho) %>%
  mutate_at(vars(-date), ~ replace_na(., 0))

# plot it 

migration_ts <- ggplot(data = willfalls_ts, aes(x = date)) +
  geom_line(aes(y = steelhead, color = "Steelhead")) +
  geom_line(aes(y = coho, color = "Coho")) +
  geom_line(aes(y = jack_coho, color = "Jack Coho")) +
  labs(title = "Migration Time Series", caption = "Counts of adult Coho, Jack Coho, and Steelhead passage in Willamette Falls",
       x = "Date",
       y = "Fish Count",
       color = "Species") +
  theme_light()

migration_ts
```

2 - 3 bullet points summarizing major patterns / notable trends in the salmon passage time series data

- Steelhead seems to have a seasonal migration/passage. 
- Steelhead populations drop in 2008 and do not seem to rebound.
- Coho and Jack Coho populations increased when Steelhead populations declined.

### Task 2 Part 2: Seasonplots

A finalized seasonplot for each species (coho, jack coho, steelhead) in its own panel. 

```{r wrangle season data}
# I was trying different ways to change the months because when i would originally plot them the month_name was not in chronological order 

willfalls_ts_season <- willfalls_ts %>%
  mutate(date = lubridate::ymd(date)) %>%
  mutate(month = lubridate::month(date)) %>%
  mutate(month_name = factor(month.abb[month], levels = month.abb)) %>%
  group_by(month_name, month) %>%
  summarize(steelhead_cnt = sum(steelhead), coho_cnt = sum(coho), jack_coho_cnt = sum(jack_coho)) %>%
  select(-month)

#plotting it 

season_plot_steelhead<- ggplot(data = willfalls_ts_season, aes(x = month_name, y = steelhead_cnt)) +
  geom_line()+
  labs(title = "Steelhead Seasonal Migration", caption = "Seasonal migration of adult Steelhead by counts through passage in Willamette Falls",
       x = "Month",
       y = "Fish Count") +
  theme_light()

season_plot_steelhead

# when I plot it with geom_line the values don't show up 

season_plot_steelhead<- ggplot(data = willfalls_ts_season, aes(x = month_name, y = steelhead_cnt)) +
  geom_point()+
  labs(title = "Steelhead Seasonal Migration", caption = "Seasonal migration of adult Steelhead by counts through passage in Willamette Falls",
       x = "Month",
       y = "Fish Count") +
  theme_light()

season_plot_steelhead

# when I use geom_point they do, put the points don't connect and I still have the issue that the months are out of order - and what do I do about the NA values? 
```

